---
title: "crop_calendar_extraction"
output: html_document
date: "2024-01-17"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(tidyverse)

```

# read crop calendar data
data was downloaded from: https://sage.nelson.wisc.edu/data-and-models/datasets/crop-calendar-dataset/

all data 

```{r calendar}
raw <- read.csv(file = 'data/raw_data/Crop_calendar_data_with_climate.csv', row.names = 'X.1')

```

Crop data for US states was extracted by: 
1) Nation code: 5
2) Crop.name.in.orignial.data:

#'CORN','SOYBEANS','WHEAT','COTTON','SORGHUM',"BARLEY","OATS","PEANUTS",
# "SUGARBEETS"
# Tobacco is not included in this file

```{r filter}

crop_day <- raw %>%
  filter(Nation.code == 5) %>% # filter by nation
  filter(Crop %in% c("Barley", "Maize", "Cotton", "Oats", "Sorghum", "Soybeans","Groundnuts",
                     "Sugarbeets", "Wheat")) %>% # filter by crop name
  #filter(Crop.name.in.original.data != "Wheat - spring")
  mutate(state_alpha = datasets::state.abb[match(Location,state.name)]) %>%
  dplyr::select(state_alpha,Crop, Crop.name.in.original.data,
                Plant.median, Harvest.median)

```

# TOBACCO was not included in the crop calendar data
1. extract crop progress data from NASS survey 

```{r tobacco}
library(rnassqs)
api_key <- as.character(read.csv("nass_key.csv", header = F)[1,1])       # api key
# Call in data via NASS API 
nassqs_auth(key = api_key)
# Specify the range of years across which you want to collect data
years <- as.list(2000:2023)  

tobacco <- plyr::ldply(years, function(x){
  
  params <- list(
    commodity_desc = "TOBACCO",
    year = x,
    agg_level_desc = "STATE", 
    source_desc = "SURVEY"  # change source to source_desc
    #domain_desc = "TOTAL"
    #statisticcat_desc = 'PROGRESS' # give error if specify this in tobacco
    #unit_desc = c("PCT TRANSPLANTED","PCT HARVESTED") # give error if specify this in tobacco
  )
  
  return(
    nassqs(params) %>%
      dplyr::select(
        commodity_desc,
        unit_desc,
        short_desc,
        year,
        state_alpha,
        state_ansi,
        begin_code,
        reference_period_desc,
        week_ending,
        Value
      ) %>%
      rename(crop_progress_pct = Value) %>%
      filter(
        grepl("PCT TRANSPLANTED|PCT HARVESTED", unit_desc) & 
          grepl("TOBACCO", commodity_desc)
      ) %>%
      
      # Filter based on specific state and short_desc conditions
      # Note: only NC has specified the progress for specific type
      filter(
        (grepl("TOBACCO, AIR-CURED LIGHT BURLEY \\(TYPE 31\\) - PROGRESS, MEASURED IN PCT TRANSPLANTED|TOBACCO, AIR-CURED LIGHT BURLEY \\(TYPE 31\\) - PROGRESS, MEASURED IN PCT HARVESTED", short_desc) & 
           state_alpha == "NC") |
          
          (grepl("TOBACCO - PROGRESS, MEASURED IN PCT TRANSPLANTED|TOBACCO - PROGRESS, MEASURED IN PCT HARVESTED", short_desc) & 
             state_alpha %in% c("KY", "TN"))
      )
  )
})


```
2. Extract the week starting date of at least 50% planted or matured 

```{r median_date}
#1) calculate the starting date of the week (week starts on Monday)
library(lubridate)

# convert character to date
tobacco$week_ending <- as.Date(tobacco$week_ending)

tobacco_prog <- tobacco %>%
  mutate(week_start = floor_date(week_ending, unit = 'week', week_start = 1)) %>%
  group_by(commodity_desc,year,state_alpha,unit_desc) %>%
  filter(crop_progress_pct >= 50) %>%
  slice_min(week_start) %>%
  ungroup() %>%
  mutate(yday = yday(week_start)) %>%
  dplyr::select(commodity_desc,year, state_alpha,unit_desc,begin_code,yday,week_start,crop_progress_pct)

tobacco_day <- tobacco_prog %>% 
  dplyr::select(commodity_desc, year, state_alpha,unit_desc,yday) %>%
  # get the median of the planting and harvest day
  group_by(state_alpha,unit_desc) %>%
  summarise(yday_median = round(median(yday))) %>%
  pivot_wider(names_from = unit_desc,values_from = yday_median) %>%
  mutate(Crop = 'Tobacco', Crop.name.in.original.data = 'Tobacco_NASS') %>%
  rename(Plant.median = `PCT TRANSPLANTED`,
         Harvest.median = `PCT HARVESTED`) %>%
  dplyr::select(state_alpha, Crop, Crop.name.in.original.data,Plant.median,Harvest.median)

crop_day_1 <- rbind.data.frame(crop_day,tobacco_day)

#write.csv(crop_day_1, file = 'data/intermediate_data/Crop_calendar_plant_harvest_day.csv')

```

# read yield county list
```{r data}

# crop yield county list
crop_county <- read.csv(file = 'data/nass_processed_yield/all_10_crops_county_list_2000_2023.csv')
# if read from excel file, the GEOID may be read as integer; convert to characters 
crop_county$GEOID <- formatC(crop_county$GEOID, width = 5, format = 'd' ,flag = '0')

```

# check if there is any missing crop calenar each state
# if data is missing, using NASS crop progress data to fill the gap

```{r crop}

crop_calendar <- crop_day_1 %>%
  mutate(Crop = case_when(
    Crop == 'Maize' ~ 'Corn',
    Crop == 'Groundnuts' ~ 'Peanuts',
    .default = Crop
  ))

library(stringr)
check_calendar <- crop_county %>%
  pivot_longer(names_to = 'crop', values_to = 'value', cols = c(BARLEY:WHEAT)) %>%
  filter(value == 1) %>%
  group_by(state_alpha, crop) %>%
  summarise(value = unique(value)) %>%
  mutate(crop = str_to_title(crop)) %>%
  left_join(crop_calendar, by = c('state_alpha', 'crop' = 'Crop'))
  
missing_data <- check_calendar %>%
  filter(is.na(Crop.name.in.original.data))

missing_data
#
#  state_alpha crop     value Crop.name.in.original.data Plant.median Harvest.median
#  <chr>       <chr>    <int> <chr>                             <dbl>          <dbl>
#1 KS          Cotton       1 NA                                   NA             NA
#2 NY          Soybeans     1 NA                                   NA             NA
#3 WV          Soybeans     1 NA                                   NA             NA

```

For the three states and crops with missing crop calendar data
Use NASS crop progress data to fill the gap

KS cotton
```{r missing}
library(rnassqs)
api_key <- as.character(read.csv("nass_key.csv", header = F)[1,1])       # api key
# Call in data via NASS API 
nassqs_auth(key = api_key)
# Specify the range of years across which you want to collect data
years <- as.list(2000:2023)  

KS_cotton <- plyr::ldply(years, function(x){
  
  params <- list(
    commodity_desc = "COTTON",
    year = x,
    agg_level_desc = "STATE", 
    source_desc = "SURVEY",  # change source to source_desc
    #domain_desc = "TOTAL"
    #state_alpha = 'KS',
    statisticcat_desc = 'PROGRESS' # give error if specify this in tobacco
    #unit_desc = c("PCT TRANSPLANTED","PCT HARVESTED") # give error if specify this in tobacco
  )
  
  return(
    nassqs(params) %>%
#  )
#})      
      
      dplyr::select(
        commodity_desc,
        unit_desc,
        short_desc,
        year,
        state_alpha,
        state_ansi,
        begin_code,
        reference_period_desc,
        week_ending,
        Value
      ) %>%
  filter(state_alpha == 'KS') %>%
  rename(crop_progress_pct = Value) %>%
  filter(
    grepl("PCT PLANTED|PCT HARVESTED", unit_desc)
  ) #%>%
)
})

# Extract the week starting date of at least 50% planted or matured 

# convert character to date
KS_cotton$week_ending <- as.Date(KS_cotton$week_ending)

KS_cotton_prog <- KS_cotton %>%
  mutate(week_start = floor_date(week_ending, unit = 'week', week_start = 1)) %>%
  group_by(commodity_desc,year,state_alpha,unit_desc) %>%
  filter(crop_progress_pct >= 50) %>%
  slice_min(week_start) %>%
  ungroup() %>%
  mutate(yday = yday(week_start)) %>%
  dplyr::select(commodity_desc,year, state_alpha,unit_desc,begin_code,yday,week_start,crop_progress_pct)

KS_cotton_day <- KS_cotton_prog %>% 
  dplyr::select(commodity_desc, year, state_alpha,unit_desc,yday) %>%
  # get the median of the planting and harvest day
  group_by(state_alpha,unit_desc) %>%
  summarise(yday_median = round(median(yday))) %>%
  pivot_wider(names_from = unit_desc,values_from = yday_median) %>%
  mutate(Crop = 'Cotton', Crop.name.in.original.data = 'Cotton_NASS') %>%
  rename(Plant.median = `PCT PLANTED`,
         Harvest.median = `PCT HARVESTED`) %>%
  dplyr::select(state_alpha, Crop, Crop.name.in.original.data,Plant.median,Harvest.median)


```

NY and WV soybeans
```{r missing}

miss_soybeans <- plyr::ldply(years, function(x){
  
  params <- list(
    commodity_desc = "SOYBEANS",
    year = x,
    agg_level_desc = "STATE", 
    source_desc = "SURVEY",  # change source to source_desc
    #domain_desc = "TOTAL"
    #state_alpha = 'KS',
    statisticcat_desc = 'PROGRESS' # give error if specify this in tobacco
    #unit_desc = c("PCT TRANSPLANTED","PCT HARVESTED") # give error if specify this in tobacco
  )
  
  return(
    nassqs(params) %>%
#  )
#})      
      
      dplyr::select(
        commodity_desc,
        unit_desc,
        short_desc,
        year,
        state_alpha,
        state_ansi,
        begin_code,
        reference_period_desc,
        week_ending,
        Value
      ) %>%
  filter(state_alpha %in% c('NY','WV')) %>%
  rename(crop_progress_pct = Value) %>%
  filter(
    grepl("PCT PLANTED|PCT HARVESTED",
          unit_desc)
  ) #%>%
)
})

# Extract the week starting date of at least 50% planted or matured 

# convert character to date
miss_soybeans$week_ending <- as.Date(miss_soybeans$week_ending)

miss_soybeans_prog <- miss_soybeans %>%
  mutate(week_start = floor_date(week_ending, unit = 'week', week_start = 1)) %>%
  group_by(commodity_desc,year,state_alpha,unit_desc) %>%
  filter(crop_progress_pct >= 50) %>%
  slice_min(week_start) %>%
  ungroup() %>%
  mutate(yday = yday(week_start)) %>%
  dplyr::select(commodity_desc,year, state_alpha,unit_desc,begin_code,yday,week_start,crop_progress_pct)

miss_soybeans_day <- miss_soybeans_prog %>% 
  dplyr::select(commodity_desc, year, state_alpha,unit_desc,yday) %>%
  # get the median of the planting and harvest day
  group_by(state_alpha,unit_desc) %>%
  summarise(yday_median = round(median(yday))) %>%
  pivot_wider(names_from = unit_desc,values_from = yday_median) %>%
  mutate(Crop = 'Soybeans', Crop.name.in.original.data = 'Soybeans_NASS') %>%
  rename(Plant.median = `PCT PLANTED`,
         Harvest.median = `PCT HARVESTED`) %>%
  dplyr::select(state_alpha, Crop, Crop.name.in.original.data,Plant.median,Harvest.median)


```

Combine results

```{r combine}
crop_day_2 <- rbind.data.frame(crop_day_1,KS_cotton_day,miss_soybeans_day)

write.csv(crop_day_2, file = 'data/intermediate_data/Crop_calendar_plant_harvest_day.csv')
```

Not RUN !
# convert day of the year to date: since the paper mentioned that this data is:"Our observations generally refer to planting and harvesting dates in the 1990s and early 2000s" (Sacks et al., 2010)

```{r date}

corn_calendar <- raw %>%
  filter(Nation.code == 5 & Crop.name.in.original.data == 'Corn' ) %>%
  mutate(Plant.median.date = as.Date(Plant.median, origin = '1999-12-31'),
         Harvest.median.date = as.Date(Harvest.median, origin = '1999-12-31'),
         state_alpha = datasets::state.abb[match(Location,state.name)]) %>%
  dplyr::select(state_alpha, Plant.median.date, Harvest.median.date) %>%
  mutate(Plant.median.month = month(Plant.median.date),
         Harvest.median.month = month(Harvest.median.date)) %>%
  mutate(Plant.median.day = day(Plant.median.date),
         Harvest.median.day = day(Harvest.median.date)) %>%
  mutate(Plant.median.month.day = paste0(sprintf("%02d", Plant.median.month), '-', sprintf("%02d",Plant.median.day)),
         Harvest.median.month.day = paste0(sprintf("%02d", Harvest.median.month), '-', sprintf("%02d", Harvest.median.day))) %>%
  select(-c(Plant.median.date,Harvest.median.date))

write.csv(corn_calendar, file = 'data/AFT-data/Drought-data/Crop_calendar/corn_planting_harvest_month_day.csv')

```

