---
title: "Check_crop_progress_missing_data"
author: "ML"
date: "2025-02-19"
output: html_document
editor_options: 
  chunk_output_type: console
---

# setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(rnassqs)
library(dplyr)
library(ggplot2)
library(tidyr)
library(reshape2)
library(caret)
library(parallel)
library(readr)
api_key <- as.character(read.csv("nass_key.csv", header = F)[1,1])       # api key

# Call in all corn yield data via NASS API ####

nassqs_auth(key = api_key)

```

# Pull data from NASS 

1. Specify years and pull crop progress data from nass 'SURVEY'

```{r data}
# Specify the range of years across which you want to collect data
years <- as.list(2000:2023)  

## Crop Progress 

# group I: 'CORN','WHEAT','SORGHUM',"BARLEY","OATS"
# group II (need script adjustment): 'COTTON',"TOBACCO",'SOYBEANS' "SUGARBEETS","PEANUTS"

# group I: can be extracted by plant and mature  
group1 <- plyr::ldply(years, function(x){
  
  params <- list(
    commodity_desc =
      c('CORN','SORGHUM',"BARLEY","OATS"),
    year = x,
    agg_level_desc = "STATE", 
    source_desc = "SURVEY",  # change source to source_desc
    #domain_desc = "TOTAL",
    statisticcat_desc = 'PROGRESS',
    unit_desc = c("PCT PLANTED", "PCT MATURE") # this description might be different for different crops
  )
  
  return(
    nassqs(params) %>%
        dplyr::select(
        commodity_desc,
        unit_desc,
        short_desc,
        year,
        state_alpha,
        state_ansi,
        begin_code,
        reference_period_desc,
        week_ending,
        Value
      ) %>%
      rename(crop_progress_pct = Value)
  )
}) 

# group II: have different definitions of plant and mature  
group2 <- plyr::ldply(years, function(x){
  
  params <- list(
    commodity_desc =
      c('COTTON','SOYBEANS','WHEAT',"TOBACCO", "SUGARBEETS","PEANUTS"),
    year = x,
    agg_level_desc = "STATE", 
    source_desc = "SURVEY",  # change source to source_desc
    #domain_desc = "TOTAL",
    statisticcat_desc = 'PROGRESS',
    unit_desc = c("PCT PLANTED", "PCT TRANSPLANTED",'PCT DROPPING LEAVES', "PCT HARVESTED", "PCT BOLLS OPENING") # this description might be different for different crops
  )
  
  return(
    nassqs(params) %>%
        dplyr::select(
        commodity_desc,
        unit_desc,
        short_desc,
        year,
        state_alpha,
        state_ansi,
        begin_code,
        reference_period_desc,
        week_ending,
        Value
      ) %>%
      rename(crop_progress_pct = Value) %>%
      # Filter the data
      filter(
      # For Soybeans-related records
        ((grepl("SOYBEANS - PROGRESS, MEASURED IN PCT PLANTED", short_desc) | 
          grepl("SOYBEANS - PROGRESS, MEASURED IN PCT DROPPING LEAVES", short_desc)) & 
          grepl("SOYBEANS", commodity_desc)) |
      
      # For Cotton-related records
        ((grepl("COTTON, UPLAND - PROGRESS, MEASURED IN PCT PLANTED", short_desc) | 
          grepl("COTTON, UPLAND - PROGRESS, MEASURED IN PCT BOLLS OPENING", short_desc)) & 
          grepl("COTTON", commodity_desc)) |
      
      # For Tobacco-related records
        ((grepl("TOBACCO, AIR-CURED LIGHT BURLEY \\(TYPE 31\\) - PROGRESS, MEASURED IN PCT TRANSPLANTED|TOBACCO, AIR-CURED LIGHT BURLEY \\(TYPE 31\\) - PROGRESS, MEASURED IN PCT HARVESTED", short_desc) & 
           state_alpha == "NC") |
          
        (grepl("TOBACCO - PROGRESS, MEASURED IN PCT TRANSPLANTED|TOBACCO - PROGRESS, MEASURED IN PCT HARVESTED", short_desc) & 
             state_alpha %in% c("KY", "TN"))) |
        
      # For Sugarbeets
          grepl("SUGARBEETS", commodity_desc) |
        
      # For Peanuts
          grepl('PEANUTS', commodity_desc) |
        
      # For Winter Wheat: Filter rows where the 'short_desc' column contains "WHEAT, WINTER"
          grepl("WHEAT, WINTER", short_desc) & grepl("WHEAT", commodity_desc)
      )
  )
})

```

2. Extract the week starting date of at least 50% planted or matured 

```{r median_date}
#1) calculate the starting date of the week (week starts on Monday)
library(lubridate)

# combine two datasets
crop <- rbind.data.frame(group1,group2)

# convert character to date
crop$week_ending <- as.Date(crop$week_ending)

corn_prog <- crop %>%
  mutate(week_start = floor_date(week_ending, unit = 'week', week_start = 1)) %>%
  group_by(commodity_desc,year,state_alpha,unit_desc) %>%
  filter(crop_progress_pct >= 50) %>%
  slice_min(week_start) %>%
  ungroup() %>%
  mutate(yday = yday(week_start)) %>%
  select(commodity_desc,year, state_alpha,unit_desc,begin_code,yday,week_start,crop_progress_pct)

```

3. Check for missing values

```{r missing}
# add week start
#corn_prog$week_start <- as.Date(corn_prog$week_start)

corn_prog_wide <- corn_prog %>% 
  select(-c(begin_code,crop_progress_pct,week_start)) %>%
  
  #  Replace all values in a column to "PCT Planted" and "PCT MATURE" 
  mutate(
    unit_desc = case_when(
      unit_desc %in% c("PCT PLANTED", "PCT TRANSPLANTED") ~ "PCT PLANTED",
      TRUE ~ "PCT MATURE")
  ) %>%
  #pivot_wider(names_from = unit_desc, values_from = c(week_start,yday)) %>%
  pivot_wider(names_from = unit_desc, values_from = yday) %>%
  rename(pct_planted = `PCT PLANTED`) %>%
  rename(pct_mature = `PCT MATURE`) %>%
  mutate(diff = case_when(pct_planted > pct_mature ~ 'check',
                          TRUE ~ 'other'))

# read the county list
crop_counties <- read.csv(file = 'data/nass_processed_yield/all_10_crops_county_list_2000_2023.csv')
# Ensure the 'GEOID' column is a string with 5 characters, left-padded with zeros
crop_counties$GEOID <- formatC(crop_counties$GEOID, width = 5, format = 'd' ,flag = '0')

crop_counties_long <- crop_counties %>%
  pivot_longer(names_to = 'crop', values_to = 'value', cols = c(BARLEY:WHEAT)) %>%
  filter(value == 1) %>%
  dplyr::select(-c(GEOID, value)) %>%
  unique()

# filter crop and states that have yield data
corn_prog_filter <- crop_counties_long %>%
  left_join(corn_prog_wide, by = c('state_alpha', 'crop' = 'commodity_desc'))

test <- corn_prog_filter %>% 
  dplyr::select(-diff) %>%
  group_by(crop,state_alpha) %>%
  summarise(across(everything(), ~ sum(!is.na(.)), .names = 'non_na_{col}')) %>%
  mutate(non_na = case_when(
    non_na_pct_planted == 0 | non_na_pct_mature == 0 ~ 0,
    TRUE ~ 1
  ))

test %>% filter(non_na == 0)

# A tibble: 22 × 6
# Groups:   crop [4]
#   crop       state_alpha non_na_year non_na_pct_planted non_na_pct_mature non_na
#   <chr>      <chr>             <int>              <int>             <int>  <dbl>
# 1 BARLEY     CA                    0                  0                 0      0
# 2 BARLEY     CO                   10                 10                 0      0
# 3 BARLEY     ID                   24                 24                 0      0
# 4 BARLEY     MN                   24                 24                 0      0
# 5 BARLEY     MT                   24                 24                 0      0
# 6 BARLEY     OR                   10                 10                 0      0
# 7 BARLEY     UT                   10                 10                 0      0
# 8 BARLEY     VA                    9                  9                 0      0
# 9 BARLEY     WA                   24                 24                 0      0
#10 CORN       CA                    0                  0                 0      0
#11 CORN       MT                   10                 10                 0      0
#12 OATS       IA                   24                 24                 0      0
#13 OATS       IL                    4                  4                 0      0
#14 OATS       MN                   24                 24                 0      0
#15 OATS       MT                   10                 10                 0      0
#16 OATS       NY                   10                 10                 0      0
#17 OATS       OH                   24                 24                 0      0
#18 OATS       OR                    9                  9                 0      0
#19 OATS       TX                   19                 19                 0      0
#20 OATS       WI                   24                 24                 0      0
#21 SUGARBEETS CA                    0                  0                 0      0
#22 SUGARBEETS NE                    0                  0                 0      0

```

4. to fix the missing value. 

Corn, Barley, and oats

```{r data}
# Specify the range of years across which you want to collect data
years <- as.list(2000:2023)  

## Crop Progress 

# group I: 'CORN','WHEAT','SORGHUM',"BARLEY","OATS"
# group II (need script adjustment): 'COTTON',"TOBACCO",'SOYBEANS' "SUGARBEETS","PEANUTS"

# group I: can be extracted by plant and mature  
barley_oat <- plyr::ldply(years, function(x){
  
  params <- list(
    commodity_desc =
      c("BARLEY","OATS"),
    year = x,
    agg_level_desc = "STATE", 
    source_desc = "SURVEY",  # change source to source_desc
    #domain_desc = "TOTAL",
    statisticcat_desc = 'PROGRESS',
    unit_desc = c("PCT MATURE", "PCT HARVESTED") # this description might be different for different crops
  )
  
  return(
    nassqs(params) %>%
        dplyr::select(
        commodity_desc,
        unit_desc,
        short_desc,
        year,
        state_alpha,
        state_ansi,
        begin_code,
        reference_period_desc,
        week_ending,
        Value
      ) %>%
      rename(crop_progress_pct = Value)
  )
})

corn <- plyr::ldply(years, function(x){
  
  params <- list(
    commodity_desc =
      c('CORN'),
    year = x,
    agg_level_desc = "STATE", 
    source_desc = "SURVEY",  # change source to source_desc
    #domain_desc = "TOTAL",
    statisticcat_desc = 'PROGRESS',
    unit_desc = c("PCT MATURE","PCT HARVESTED") # this description might be different for different crops
  )
  
  return(
    nassqs(params) %>%
        dplyr::select(
        commodity_desc,
        unit_desc,
        short_desc,
        year,
        state_alpha,
        state_ansi,
        begin_code,
        reference_period_desc,
        week_ending,
        Value
      ) %>%
      rename(crop_progress_pct = Value)
  )
}) %>%
  filter(
      # For Soybeans-related records
          grepl("CORN - PROGRESS, MEASURED IN PCT MATURE", short_desc) |
          grepl("CORN, GRAIN - PROGRESS, MEASURED IN PCT HARVESTED", short_desc)
  )
```

2. Extract the week starting date of harvest and mature 

```{r median_date}
#1) calculate the starting date of the week (week starts on Monday)
library(lubridate)
library(ggplot2)

# convert character to date
barley_oat$begin_code <- as.numeric(barley_oat$begin_code)

barley_oat_prog <- barley_oat %>%
  dplyr::select(-c(short_desc,state_ansi,reference_period_desc,week_ending)) %>%
  group_by(commodity_desc,state_alpha,year) %>%
  mutate(rel_begin_code = begin_code - min(begin_code)+1)

# barley 
barley_prog <- barley_oat_prog %>% 
  filter(commodity_desc == 'BARLEY')

ggplot(data = barley_prog) +
  geom_point(aes(x = rel_begin_code, y = crop_progress_pct, color = unit_desc)) +
  geom_smooth(aes(x = rel_begin_code, y = crop_progress_pct, color = unit_desc), method = 'loess') +
  facet_wrap(~state_alpha) +
  labs(title = 'Barley')

# Oats
oat_prog <- barley_oat_prog %>% 
  filter(commodity_desc == 'OATS')

ggplot(data = oat_prog) +
  geom_point(aes(x = rel_begin_code, y = crop_progress_pct, color = unit_desc)) +
  geom_smooth(aes(x = rel_begin_code, y = crop_progress_pct, color = unit_desc), method = 'loess') +
  facet_wrap(~state_alpha) +
  labs(title = 'Oats')

# corn
# convert character to date
corn$begin_code <- as.numeric(corn$begin_code)

corn_prog <- corn %>%
  dplyr::select(-c(short_desc,state_ansi,reference_period_desc,week_ending)) %>%
  group_by(commodity_desc,state_alpha,year) %>%
  mutate(rel_begin_code = begin_code - min(begin_code)+1)

ggplot(data = corn_prog) +
  geom_point(aes(x = rel_begin_code, y = crop_progress_pct, color = unit_desc)) +
  geom_smooth(aes(x = rel_begin_code, y = crop_progress_pct, color = unit_desc), method = 'loess') +
  facet_wrap(~state_alpha) +
  labs(title = 'Corn')

```

Results: A few states are missing the mature dates, but have harvest dates for barley and oats

For barley and oats: when extracting the mature dates, >50% of mature is usually overlaps with 25% harvest. So will use >= 25% harvest date for mature dates

For corn: when extracting the mature dates, >50% of mature is usually overlaps with 5~10% harvest. So will use >= 7.5% harvest date for mature dates
