---
title: "crop_progress"
author: "ML"
date: "2025-01-16"
output: html_document
editor_options: 
  chunk_output_type: console
---

# setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(rnassqs)
library(dplyr)
library(ggplot2)
library(tidyr)
library(reshape2)
library(caret)
library(parallel)
library(readr)
api_key <- as.character(read.csv("nass_key.csv", header = F)[1,1])       # api key

# Call in all corn yield data via NASS API ####

nassqs_auth(key = api_key)

```

# Pull data from NASS 

1. Specify years and pull crop progress data from nass 'SURVEY'

Results: A few states are missing the mature dates, but have harvest dates for barley and oats

For barley: When mature data is not available, use >= 25% harvest date to substitute mature dates
Barley states: "CA" "CO" "ID" "MN" "MT" "OR" "UT" "VA" "WA"

For oats: When mature data is not available, use >= 25% harvest date to substitute mature dates
Oats states: "IA" "IL" "MN" "MT" "NY" "OH" "OR" "TX" "WI"

For corn: When mature data is not available, use >= 7.5% harvest date for mature dates
Corn states: 'MT'

```{r data}
# Specify the range of years across which you want to collect data
years <- as.list(2000:2023)  

## Crop Progress 

# group I: 'CORN','WHEAT','SORGHUM',"BARLEY","OATS"
# group II (need script adjustment): 'COTTON',"TOBACCO",'SOYBEANS' "SUGARBEETS","PEANUTS"

# group I: can be extracted by plant and mature  
group1 <- plyr::ldply(years, function(x){
  
  params <- list(
    commodity_desc =
      c('CORN','SORGHUM',"BARLEY","OATS"),
    year = x,
    agg_level_desc = "STATE", 
    source_desc = "SURVEY",  # change source to source_desc
    #domain_desc = "TOTAL",
    statisticcat_desc = 'PROGRESS',
    unit_desc = c("PCT PLANTED", "PCT MATURE","PCT HARVESTED") # this description might be different for different crops
  )
  
  return(
    nassqs(params) %>%
        dplyr::select(
        commodity_desc,
        unit_desc,
        short_desc,
        year,
        state_alpha,
        state_ansi,
        begin_code,
        reference_period_desc,
        week_ending,
        Value
      ) %>%
      rename(crop_progress_pct = Value)
  )
}) %>%
  filter(
      # For Corn-related records
        (grepl("CORN - PROGRESS, MEASURED IN PCT PLANTED", short_desc) | 
          grepl("CORN - PROGRESS, MEASURED IN PCT MATURE", short_desc) |
          (grepl("CORN, GRAIN - PROGRESS, MEASURED IN PCT HARVESTED", short_desc) & state_alpha == "MT" )) |
          
       # For barley
         (grepl("BARLEY - PROGRESS, MEASURED IN PCT PLANTED", short_desc) | 
          grepl("BARLEY - PROGRESS, MEASURED IN PCT MATURE", short_desc) |
          (grepl("BARLEY - PROGRESS, MEASURED IN PCT HARVESTED", short_desc) & 
             state_alpha %in% c("CA", "CO", "ID", "MN", "MT", "OR", "UT", "VA", "WA"))) |
        
       # For oats
         (grepl("OATS - PROGRESS, MEASURED IN PCT PLANTED", short_desc) | 
          grepl("OATS - PROGRESS, MEASURED IN PCT MATURE", short_desc) |
          (grepl("OATS - PROGRESS, MEASURED IN PCT HARVESTED", short_desc) & 
             state_alpha %in% c("IA", "IL", "MN", "MT", "NY", "OH", "OR", "TX", "WI"))) |
        
       # For sorghum
        (grepl("SORGHUM - PROGRESS, MEASURED IN PCT PLANTED", short_desc) | 
          grepl("SORGHUM - PROGRESS, MEASURED IN PCT MATURE", short_desc))
        
  )

# group II: have different definitions of plant and mature  
group2 <- plyr::ldply(years, function(x){
  
  params <- list(
    commodity_desc =
      c('COTTON','SOYBEANS','WHEAT',"TOBACCO", "SUGARBEETS","PEANUTS"),
    year = x,
    agg_level_desc = "STATE", 
    source_desc = "SURVEY",  # change source to source_desc
    #domain_desc = "TOTAL",
    statisticcat_desc = 'PROGRESS',
    unit_desc = c("PCT PLANTED", "PCT TRANSPLANTED",'PCT DROPPING LEAVES', "PCT HARVESTED", "PCT BOLLS OPENING") # this description might be different for different crops
  )
  
  return(
    nassqs(params) %>%
        dplyr::select(
        commodity_desc,
        unit_desc,
        short_desc,
        year,
        state_alpha,
        state_ansi,
        begin_code,
        reference_period_desc,
        week_ending,
        Value
      ) %>%
      rename(crop_progress_pct = Value) %>%
      # Filter the data
      filter(
      # For Soybeans-related records
        ((grepl("SOYBEANS - PROGRESS, MEASURED IN PCT PLANTED", short_desc) | 
          grepl("SOYBEANS - PROGRESS, MEASURED IN PCT DROPPING LEAVES", short_desc)) & 
          grepl("SOYBEANS", commodity_desc)) |
      
      # For Cotton-related records
        ((grepl("COTTON, UPLAND - PROGRESS, MEASURED IN PCT PLANTED", short_desc) | 
          grepl("COTTON, UPLAND - PROGRESS, MEASURED IN PCT BOLLS OPENING", short_desc)) & 
          grepl("COTTON", commodity_desc)) |
      
      # For Tobacco-related records
        ((grepl("TOBACCO, AIR-CURED LIGHT BURLEY \\(TYPE 31\\) - PROGRESS, MEASURED IN PCT TRANSPLANTED|TOBACCO, AIR-CURED LIGHT BURLEY \\(TYPE 31\\) - PROGRESS, MEASURED IN PCT HARVESTED", short_desc) & 
           state_alpha == "NC") |
          
        (grepl("TOBACCO - PROGRESS, MEASURED IN PCT TRANSPLANTED|TOBACCO - PROGRESS, MEASURED IN PCT HARVESTED", short_desc) & 
             state_alpha %in% c("KY", "TN"))) |
        
      # For Sugarbeets
          grepl("SUGARBEETS", commodity_desc) |
        
      # For Peanuts
          grepl('PEANUTS', commodity_desc) |
        
      # For Winter Wheat: Filter rows where the 'short_desc' column contains "WHEAT, WINTER"
          grepl("WHEAT, WINTER", short_desc) & grepl("WHEAT", commodity_desc)
      )
  )
})

```

2. Extract the week starting date of at least 50% planted or matured 

Note from the missing results: 

Results: A few states are missing the mature dates, but have harvest dates for barley and oats

For barley: When mature data is not available, use >= 25% harvest date to substitute mature dates
Barley states: "CA" "CO" "ID" "MN" "MT" "OR" "UT" "VA" "WA"

For oats: When mature data is not available, use >= 25% harvest date to substitute mature dates
Oats states: "IA" "IL" "MN" "MT" "NY" "OH" "OR" "TX" "WI"

For corn: When mature data is not available, use >= 7.5% harvest date for mature dates
Corn states: 'MT'

```{r median_date}
#1) calculate the starting date of the week (week starts on Monday)
library(lubridate)

# combine two datasets
crop <- rbind.data.frame(group1,group2)

# convert character to date
crop$week_ending <- as.Date(crop$week_ending)

corn_prog <- crop %>%
  mutate(week_start = floor_date(week_ending, unit = 'week', week_start = 1)) %>%
  group_by(commodity_desc,year,state_alpha,unit_desc) %>%
  filter( 
   # If BARLEY..., keep rows with crop_progress_pct >= 25
    (short_desc == "BARLEY - PROGRESS, MEASURED IN PCT HARVESTED" & 
       (crop_progress_pct >= 25 & crop_progress_pct < 100)) |
   # If OATS..., keep rows with crop_progress_pct >= 25
    (short_desc == "OATS - PROGRESS, MEASURED IN PCT HARVESTED" & 
       (crop_progress_pct >= 25 & crop_progress_pct < 100)) | 
   # If CORN..., keep rows with crop_progress_pct >= 7.5
    (short_desc == "CORN, GRAIN - PROGRESS, MEASURED IN PCT HARVESTED" & 
       (crop_progress_pct >= 7.5 & crop_progress_pct < 100)) |
   # For everything else, keep rows with crop_progress_pct >= 50
    (!short_desc %in% c("BARLEY - PROGRESS, MEASURED IN PCT HARVESTED",
                         "OATS - PROGRESS, MEASURED IN PCT HARVESTED",
                         "CORN, GRAIN - PROGRESS, MEASURED IN PCT HARVESTED") &
       (crop_progress_pct >= 50 & crop_progress_pct < 100)) 
  ) %>%
  slice_min(week_start) %>%
  ungroup() %>%
  mutate(yday = yday(week_start)) %>%
  select(commodity_desc,year, state_alpha,unit_desc,begin_code,yday,week_start,crop_progress_pct)

```

3. Check for missing values

```{r missing}
# add week start
#corn_prog$week_start <- as.Date(corn_prog$week_start)

corn_prog_wide <- corn_prog %>% 
  select(-c(begin_code,crop_progress_pct,week_start)) %>%
  
  #  Replace all values in a column to "PCT Planted" and "PCT MATURE" 
  mutate(
    unit_desc = case_when(
      unit_desc %in% c("PCT PLANTED", "PCT TRANSPLANTED") ~ "planted",
      TRUE ~ "mature")
  ) %>%
  pivot_wider(names_from = unit_desc, values_from = yday) 

# read the county list
crop_counties <- read.csv(file = 'data/nass_processed_yield/all_10_crops_county_list_2000_2023.csv')
# Ensure the 'GEOID' column is a string with 5 characters, left-padded with zeros
crop_counties$GEOID <- formatC(crop_counties$GEOID, width = 5, format = 'd' ,flag = '0')

crop_counties_long <- crop_counties %>%
  pivot_longer(names_to = 'crop', values_to = 'value', cols = c(BARLEY:WHEAT)) %>%
  filter(value == 1) %>%
  dplyr::select(-c(GEOID, value)) %>%
  unique()

# filter crop and states that have yield data
corn_prog_filter <- crop_counties_long %>%
  left_join(corn_prog_wide, by = c('state_alpha', 'crop' = 'commodity_desc'))

test <- corn_prog_filter %>% 
  group_by(crop,state_alpha) %>%
  summarise(across(everything(), ~ sum(!is.na(.)), .names = 'non_na_{col}')) %>%
  mutate(non_na = case_when(
    non_na_planted == 0 | non_na_mature == 0 ~ 0,
    TRUE ~ 1
  ))

test %>% filter(non_na == 0)

# A tibble: 4 × 6
# Groups:   crop [3]
#  crop       state_alpha non_na_year non_na_mature non_na_planted non_na
#  <chr>      <chr>             <int>         <int>          <int>  <dbl>
#1 BARLEY     CA                    0             0              0      0
#2 CORN       CA                    0             0              0      0
#3 SUGARBEETS CA                    0             0              0      0
#4 SUGARBEETS NE                    0             0              0      0

```


4. calculate the state average across years

```{r average}
state_avg <- corn_prog_filter %>%
  group_by(crop,state_alpha) %>%
  summarise(planted = mean(planted, na.rm = TRUE),
            mature = mean(mature, na.rm = TRUE)) %>%
  as.data.frame()

#corn_growing_length <- state_avg %>%
#  filter(crop == 'CORN') %>%
#  mutate(length = mature - planted) %>%
#  summarise(mean_length = mean(length, na.rm = TRUE)) 
#130.4486

#sugarbeets_growing_length <- state_avg %>%
#  filter(crop == 'SUGARBEETS') %>%
#  mutate(length = mature - planted) %>%
#  summarise(mean_length = mean(length, na.rm = TRUE)) 
#172.375

# there are 4 rows with no data at all,
# try to consult the crop caldenar data
  
  # For CA barley,the only barley county in CA is 06079 (San Luis Obispo)
  # use winter barley planted date from the crop calendar data 
  # farmers usually harvest 2-3 weeks or more after maturity
  # so use the harvest date from the crop calendar data minus 21 days
state_avg[2, c('planted','mature')] <- c(6.5, 181 - 21)
 # For CA corn: use planted date from the crop calendar data
state_avg[17, c('planted','mature')] <- c(135,135+130.4)
# For CA sugarbeets: use planted date from the crop calendar data
state_avg[130, c('planted','mature')] <- c(265,(172.4 - (365-265)))
# For NE sugarbeets: use planted date from the crop calendar data
state_avg[137, c('planted','mature')] <- c(108,108+172.4)
  
```

5. Save the crop progress data

```{r save}

saveRDS(state_avg, file = 'data/intermediate_data/Crop_progress_plant_mature.rds')

```