---
title: "prism_growing_season_extraction_mean_for_all_crops"
author: "ML"
date: "2025-01-31"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(purrr)
library(dplyr)
library(stringr)
```

# read crop calendar data and yield county list
```{r data}

# crop calendar data
crop_calendar <- read.csv(file = 'data/intermediate_data/Crop_calendar_plant_harvest_day.csv', row.names = 'X')

# rename Maize to corn and Groundnuts to peanuts
crop_calendar <- crop_calendar %>%
  mutate(Crop = case_when(
    Crop == 'Maize' ~ 'Corn',
    Crop == 'Groundnuts' ~ 'Peanuts',
    TRUE ~ Crop
  )) %>%
  mutate(Crop = str_to_upper(Crop)) # convert crop name to upper case type

# crop yield county list
crop_county <- read.csv(file = 'data/nass_processed_yield/all_10_crops_county_list_2000_2023.csv')
# if read from excel file, the GEOID may be read as integer; convert to characters 
crop_county$GEOID <- formatC(crop_county$GEOID, width = 5, format = 'd' ,flag = '0')

```

# check if crop calenar is unique for each state
Because crops like barley and oats have winter and spring for one state

```{r crop}
library(stringr)
check_calendar <- crop_county %>%
  pivot_longer(names_to = 'crop', values_to = 'value', cols = c(BARLEY:WHEAT)) %>%
  filter(value == 1) %>%
  group_by(state_alpha, crop) %>%
  summarise(value = unique(value)) %>%
  mutate(crop = str_to_title(crop)) %>% # convert crop name to title type
  left_join(crop_calendar, by = c('state_alpha', 'crop' = 'Crop')) %>% 
  filter(Crop.name.in.original.data != "Wheat - spring") %>%# filter out spring wheat 
  group_by(state_alpha, crop) %>%
  summarise(n = sum(value))

# only CA contains both winter and spring Barley
# checked the crop_county file, the only barley county in CA is 06079 (San Luis Obispo)
# searched the internet: https://www.slocounty.ca.gov/departments/agriculture-weights-and-measures/department-news/2021-annual-crop-statistics-released
# "Although prices were high for both grain hay and barley, 
# yields fell significantly since those crops are dependent on winter rainfall and not supplemental irrigation."
# Therefor, Barley should be a winter crop in this county. So filter out the CA Barley-spring option
```

# update calendar data based on the check results 

```{r update}
crop_calendar_1 <- crop_county %>%
  pivot_longer(names_to = 'crop', values_to = 'value', cols = c(BARLEY:WHEAT)) %>%
  filter(value == 1) %>%
  left_join(crop_calendar, by = c('state_alpha', 'crop' = 'Crop')) %>% 
  filter(Crop.name.in.original.data != "Wheat - spring") %>%
  filter(!(Crop.name.in.original.data == 'Barley - spring' & state_alpha == 'CA')) %>%
  select(-value)

```

# check crop growing season

```{r season}

check_season <- crop_county %>%
  pivot_longer(names_to = 'crop', values_to = 'value', cols = c(BARLEY:WHEAT)) %>%
  filter(value == 1) %>%
  group_by(state_alpha, crop) %>%
  summarise(value = unique(value)) %>%
  mutate(crop = str_to_title(crop)) %>% # convert crop name to title type
  left_join(crop_calendar, by = c('state_alpha', 'crop' = 'Crop')) %>% 
  filter(Crop.name.in.original.data != "Wheat - spring") %>%# filter out spring wheat 
  filter(!(Crop.name.in.original.data == 'Barley - spring' & state_alpha == 'CA')) %>%
  mutate(season = case_when(Plant.median > Harvest.median ~ 'winter',
                            Plant.median < Harvest.median ~ 'spring',
                            TRUE ~ 'other')) %>%
  mutate(season_length = case_when(
    Plant.median > Harvest.median ~ (365-Plant.median) + Harvest.median,
    Plant.median < Harvest.median ~ Harvest.median - Plant.median,
                            TRUE ~ NA_real_))

```

# calculate cumulative precipitation during the growing season (plant-harvest using the crop calendar data)

1. Same-Year Planting and Harvesting: When the planting median (Plant.median) is less than the harvest median (Harvest.median), we sum the county_mean values between these two dates within the same year.

2. Cross-Year Planting and Harvesting: When planting occurs in the previous year (Plant.median > Harvest.median), we:

Sum county_mean values from the planting date to the end of the previous year.

Sum county_mean values from the start of the current year to the harvest date.

3. Calculate the total ppt for each year, and then get the yearly mean for the 2000-2023 period

```{r ppt}
#library(lubridate)
# cumulative ppt
ppt <- readRDS(file = 'data/prism_daily_county_mean_1999-2023/ppt_prism_daily_county_mean_1999-2023.rds')
names(ppt)

# convert date to Date format
ppt$date <- as.Date(as.character(ppt$date), format = '%Y%m%d')

ppt <- ppt %>% rename(state_alpha = STUSPS) 

# if read from excel file, the GEOID may be read as integer; convert to characters 
#ppt$GEOID <- formatC(ppt$GEOID, width = 5, format = 'd' ,flag = '0')

# get the year and day of the year for each date
ppt$year <- lubridate::year(ppt$date)
ppt$yday <- lubridate::yday(ppt$date)

# Define function to compute cumulative precipitation for a single crop
calc_cum_ppt <- function(crop_name) {
  
  # Filter crop calendar for the current crop and relevant columns
  calendar = crop_calendar_1 %>%
    filter(crop == crop_name) %>%
    distinct(state_alpha,Plant.median,Harvest.median)
    
  # Join with precipitation data for the relevant GEOIDs and years
  cum_ppt = ppt %>%
    inner_join(calendar, by = 'state_alpha') %>% 
    # filter by crop counties
    filter(GEOID %in% crop_county$GEOID[crop_county[[crop_name]]==1])
  
  # Case 1: Plant and harvest occur within the same year
  same_year_sum <- cum_ppt %>%
    group_by(GEOID,state_alpha, year) %>%
    filter(Plant.median <= Harvest.median & year > 1999) %>%
    summarise(
      total_sum = sum(
        county_mean[yday >= Plant.median & yday <= Harvest.median],
        na.rm = TRUE
      ),
      .groups = "drop"
    )
  
  # Case 2: Plant occurs in previous year, harvest in current year
  cross_year <- cum_ppt %>%
    select(-c(date,stable,type,NAME)) %>%
    group_by(GEOID,state_alpha, year) %>%
    filter(Plant.median > Harvest.median)
    # Join with previous year's data for the same GEOID and crop
  
  cross_year_sum <- cross_year %>%
    # Calculate contributions from previous and current years
    group_by(GEOID, state_alpha, year) %>%
    summarise(
      prev_sum = sum(
        county_mean[yday >= Plant.median],  # From planting to end of prev year
        na.rm = TRUE
      ),
      current_sum = sum(
        county_mean[yday <= Harvest.median],  # From start to harvest in current year
        na.rm = TRUE
      )
    ) %>%
    group_by(GEOID, state_alpha) %>%
    arrange(year) %>% # Ensure data is ordered by year within each group
    mutate(
      total_sum = lag(prev_sum, default = 0) + current_sum  # Previous year's prev_sum + current_sum
    ) %>%
    ungroup() %>%
    dplyr::select(-c(prev_sum,current_sum)) %>%
    filter(year != 1999) # don't include 1999
  
  # Combine results and add crop identifier
  bind_rows(same_year_sum, cross_year_sum) %>%
    group_by(state_alpha, GEOID) %>% 
    mutate(cum_ppt_mean = mean(total_sum)) %>% # calculate the yearly mean for each GEOID
    mutate(crop = crop_name) %>%
    rename(annual_cum_ppt = total_sum)
  
}

# process all crops

all_crops <- unique(crop_calendar_1$crop)
final_result <- lapply(all_crops, calc_cum_ppt) %>% bind_rows()

saveRDS(final_result, file = 'data/weather_results/crop_county_mean_cumulative_ppt_2000_2023.rds')

#final_result <- readRDS(file = 'data/weather_results/crop_county_mean_cumulative_ppt_2000_2023.rds')

# check data distribution by states
#ggplot(final_result, aes(x = cum_ppt_mean)) +
#    geom_histogram(
#        aes(y = after_stat(density)), 
#        binwidth = 10,  # Adjust based on your data range
#        fill = "steelblue", 
#        color = "black"
#    ) +
#    facet_wrap(~state_alpha, scales = "free_y")

# check data distribution by crops
#ggplot(final_result, aes(x = cum_ppt_mean)) +
#    geom_histogram(
#        aes(y = after_stat(density)), 
#        binwidth = 10,  # Adjust based on your data range
#        fill = "steelblue", 
#        color = "black"
#    ) +
#    facet_wrap(~crop, scales = "free_y")

```

# calculate mean tmax, tmin, tmean during the growing season (plant-harvest using the crop calendar data)

```{r mean}

# loop through three temp types and calculate means

type <- c('tmax','tmin','tmean')

for (i in type) {
  
 t <- readRDS(file = paste0('data/prism_daily_county_mean_1999-2023/',i,'_prism_daily_county_mean_1999-2023.rds'))

# convert date to Date format
 t$date <- as.Date(as.character(t$date), format = '%Y%m%d')

 # if read from CSV file, make sure the GEOID is converted to a charactor
 #t$GEOID <- formatC(t$GEOID, width = 5, format = 'd' ,flag = '0')

 # get the year and day of the year for each date
 
 t <- t %>%
   rename(state_alpha = STUSPS) %>%
   mutate(year = lubridate::year(date),
          yday = lubridate::yday(date)
    )

 # Define temperature calculation function
  calc_temp <- function(crop_name) {
  
  # Filter crop calendar for the current crop and relevant columns
    calendar = crop_calendar_1 %>%
      filter(crop == crop_name) %>%
      distinct(state_alpha,Plant.median,Harvest.median)
    
  # Join with precipitation data for the relevant GEOIDs and years
    mean_t = t %>%
      inner_join(calendar, by = 'state_alpha') %>% 
      # filter by crop counties
      filter(GEOID %in% crop_county$GEOID[crop_county[[crop_name]]==1])
  
  # Case 1: Plant and harvest occur within the same year
  same_year_mean <- mean_t %>%
    group_by(GEOID,state_alpha, year) %>%
    filter(Plant.median <= Harvest.median & year > 1999) %>% # don't include 1999
    summarise(
      year_mean = mean(
        county_mean[yday >= Plant.median & yday <= Harvest.median],
        na.rm = TRUE
      ),
      .groups = "drop"
    )
  
  # Case 2: Plant occurs in previous year, harvest in current year
  cross_year <- mean_t %>%
    select(-c(date,stable,type,NAME)) %>%
    group_by(GEOID,state_alpha, year) %>%
    filter(Plant.median > Harvest.median)
    # Join with previous year's data for the same GEOID and crop
  
  cross_year_mean <- cross_year %>%
    # Calculate contributions from previous and current years
    group_by(GEOID, state_alpha, year) %>%
    summarise(
      prev_sum = sum(
        county_mean[yday >= Plant.median],  # From planting to end of prev year
        na.rm = TRUE
      ),
      
      prev_n = sum(yday >= Plant.median, na.rm = TRUE), # calculate the number of days 
      
      current_sum = sum(
        county_mean[yday <= Harvest.median],  # From start of the year to harvest in current year
        na.rm = TRUE
      ),
      
      current_n = sum(yday <= Harvest.median, na.rm = TRUE) # calculate the number of days
      
    ) %>%
    
    group_by(GEOID, state_alpha) %>%
    arrange(year) %>% # Ensure data is ordered by year within each group
    mutate(
      # For each year, add the previous year's sum/count (using lag)
      total_sum = lag(prev_sum, default = 0) + current_sum,  # Previous year's prev_sum + current_sum
      total_n = lag(prev_n, default = 0) + current_n, # total number of days
      year_mean = total_sum/total_n
    ) %>%
    ungroup() %>%
    dplyr::select(-c(prev_sum,current_sum, prev_n, current_n,total_sum, total_n)) %>%
    filter(year != 1999) # don't include 1999
  
  # Combine results and add crop identifier
  test <- bind_rows(same_year_mean, cross_year_mean) %>%
    group_by(state_alpha, GEOID) %>% 
    # calculate mean and name dynamically based on the type
    mutate(!!paste0('mean_',i) := mean(year_mean)) %>% 
    mutate(crop = crop_name) %>%
    rename(!!paste0('annual_',i) := year_mean)
  
  }

  # process all crops
  all_crops <- unique(crop_calendar_1$crop)
  final_result <- lapply(all_crops, calc_temp) %>% bind_rows()

  saveRDS(final_result, file = paste0('data/weather_results/crop_county_mean_', i, '_2000_2023.rds'))

}

# check the data distribution 
#tmax <- readRDS('data/weather_results/crop_county_mean_tmax_2000_2023.rds')
#tmin <- readRDS('data/weather_results/crop_county_mean_tmin_2000_2023.rds')
# check data distribution by states
#ggplot(tmin, aes(x = mean_tmin)) +
#    geom_histogram(
#        aes(y = after_stat(density)), 
#        binwidth = 10,  # Adjust based on your data range
#        fill = "steelblue", 
#        color = "black"
#    ) +
#    facet_wrap(~state_alpha, scales = "free_y")

# check data distribution by crops
#ggplot(tmin, aes(x = mean_tmin)) +
#    geom_histogram(
#        aes(y = after_stat(density)), 
#        binwidth = 10,  # Adjust based on your data range
#        fill = "steelblue", 
#        color = "black"
#    ) +
#    facet_wrap(~crop, scales = "free_y")

#cor.test(tmax$mean_tmax,final_result$mean_tmean) # 0.98
#cor.test(tmin$mean_tmin,final_result$mean_tmean) # 0.99

# Note: mean_tmean is highly correlated with mean_tmax, and mean_tmin
# Use tmean in the model should be good.

```

# calculate GDD: using the (plant-mature) crop progress data
Different crop have different thresholds: 
Corn, soybeans, and wheat (Schillerberg et al., 2023 https://doi.org/10.1016/j.agrformet.2023.109620)
Cotton: (Viator et al., 2005 https://acsess.onlinelibrary.wiley.com/doi/10.2134/agronj2005.0494)
Barley, oats:
Peanut: (Oakes et al., 2020 https://doi.org/10.3146/PS20-16.1)
Sugerbeets: (Bradley et al., 2017 DOI:  10.5274/jsbr.54.1.60)

Corn:         t_base 10; t_upper = 30 C 
Soybeans:     t_base 10; t_upper = 30 C
Sorghum:      t_base 10; t_upper = 30 C
Sugerbeets:   t_base 10; t_upper = 30 C
winter wheat: t_base 0 ; t_upper = 30 C
Barley:       t_base 0 ; t_upper = 30 C 
Oats:         t_base 0 ; t_upper = 30 C
Cotton:       t_base 15.5; t_upper = 30 C
Peanut:       t_base 13; t_upper = 30 C
Tobacco:      t_base 10; t_upper = 32.2 C


```{r gdd}
library(stringr)

# read crop progress data
crop_progress <- readRDS(file = 'data/Crop_progress_plant_mature.rds')

#crop_progress$crop <- str_to_title(crop_progress$crop)

# read GDD temperature threshold
gdd_temp <- read.csv(file = 'data/GDD_temperature_threshold.csv')

# read tmax and tmin
tmax <- readRDS(file = 'data/prism_daily_county_mean_1999-2023/tmax_prism_daily_county_mean_1999-2023.rds')
tmax <- tmax %>% 
  select(date,GEOID,STUSPS,county_mean) %>%
  rename(tmax = county_mean)

tmin <- readRDS(file = 'data/prism_daily_county_mean_1999-2023/tmin_prism_daily_county_mean_1999-2023.rds')
tmin <- tmin %>% 
  select(date,GEOID,STUSPS,county_mean) %>%
  rename(tmin = county_mean)

#tmean <- readRDS(file = 'data/prism_daily_county_mean_1999-2023/tmean_prism_daily_county_mean_1999-2023.rds')
#tmin <- tmin %>% 
#  select(date,GEOID,STUSPS,county_mean) %>%
#  rename(tmin = county_mean)

tmax_tmin <- tmax %>%
  inner_join(tmin, by = c('date', 'GEOID','STUSPS')) %>%
  rename(state_alpha = STUSPS)

tmax_tmin$date <- as.Date(as.character(tmax_tmin$date), format = '%Y%m%d')
#tmax_tmin$GEOID <- formatC(tmax_tmin$GEOID, width = 5, format = 'd' ,flag = '0')

tmax_tmin <- tmax_tmin %>%
   mutate(year = lubridate::year(date),
          yday = lubridate::yday(date)
    )

# Calculate GDD 
# if t_base = 10, t_upper = 30

#1) GDD = 0, if tmax > tmin > 30 
#2) GDD = (30 + tmin)/2 - 10, if tmax > 30 > tmin > 10
#3) GDD = (tmax + tmin)2 - 10 , if 30 > tmax > tmin > 10
#4) GDD = (tmax + 10)/2 - 10, if 30 > tmax > 10 > tmin
#5) GDD = 0, if 10 > tmax
#6) GDD = (30+10)/2 - 10, if tmax > 30 > 10 > tmin

calc_gdd <- function(crop_name,t_base,t_upper) {
  
  # Filter crop calendar for the current crop and relevant columns
    calendar = crop_progress %>%
      filter(crop == crop_name) %>%
      distinct(state_alpha,planted, mature)
    
  # Join with precipitation data for the relevant GEOIDs and years
    temp = tmax_tmin %>%
      inner_join(calendar, by = 'state_alpha') %>%
      # filter by crop counties
      filter(GEOID %in% crop_county$GEOID[crop_county[[crop_name]]==1])
  
  # Case 1: Plant and harvest occur within the same year
  same_year_mean <- temp %>%
    group_by(GEOID,state_alpha, year) %>%
    filter(planted <= mature & year > 1999) %>% # don't include 1999
    mutate(
      GDD_day = case_when(tmax > t_upper & tmin > t_upper ~ 0,
                      tmax > t_upper & tmin <= t_upper & tmin > t_base ~ (t_upper + tmin)/2 - t_base,
                      tmax <= t_upper & tmax > t_base & tmin > t_base ~ (tmax + tmin)/2 - t_base,
                      tmax <= t_upper & tmax > t_base & tmin <= t_base ~ (tmax + t_base)/2 - t_base,
                      tmax <= t_base ~ 0,
                      tmax > t_upper & tmin <= t_base ~ (t_upper + t_base)/2 - t_base,
                      TRUE ~ NA_real_)
    ) %>%
    group_by(year, state_alpha, GEOID) %>%
    summarise(GDD = sum(
      GDD_day[yday >= planted & yday <= mature], 
      na.rm = TRUE), 
      .groups = 'drop')
  
  # Case 2: Plant occurs in previous year, harvest in current year
  cross_year <- temp %>%
    select(-date) %>%
    group_by(GEOID,state_alpha, year) %>%
    filter(planted > mature)
  
  cross_year_mean <- cross_year %>%
    # Calculate contributions from previous and current years
    group_by(GEOID, state_alpha, year) %>%
    mutate(
      GDD_day = case_when(tmax > t_upper & tmin > t_upper ~ 0,
                      tmax > t_upper & tmin <= t_upper & tmin > t_base ~ (t_upper + tmin)/2 - t_base,
                      tmax <= t_upper & tmax > t_base & tmin > t_base ~ (tmax + tmin)/2 - t_base,
                      tmax <= t_upper & tmax > t_base & tmin <= t_base ~ (tmax + t_base)/2 - t_base,
                      tmax <= t_base ~ 0,
                      tmax > t_upper & tmin <= t_base ~ (t_upper + t_base)/2 - t_base,
                      TRUE ~ NA_real_)
    ) %>%
    group_by(year, state_alpha, GEOID) %>%
    summarise(
      prev_sum = sum(
        GDD_day[yday >= planted],  # From planting to end of prev year
        na.rm = TRUE
      ),
      
      current_sum = sum(
        GDD_day[yday <= mature],  # From start of the year to harvest in current year
        na.rm = TRUE
      ),
      .groups = 'drop'
      
    ) %>%
    group_by(GEOID, state_alpha) %>%
    arrange(year) %>% # Ensure data is ordered by year within each group
    mutate(
      # For each year, add the previous year's sum/count (using lag)
      GDD = lag(prev_sum, default = 0) + current_sum
    ) %>%
    ungroup() %>%
    dplyr::select(-c(prev_sum,current_sum)) %>%
    filter(year != 1999) # don't include 1999
  
  # Combine results and add crop identifier
  test <- bind_rows(same_year_mean, cross_year_mean) %>%
    group_by(state_alpha, GEOID) %>% 
    mutate(mean_GDD = mean(GDD, na.rm = TRUE)) %>% 
    mutate(crop = crop_name) %>%
    rename(annual_GDD = GDD)
  
}

  # process all crops
  final_result <- mapply(
    FUN = calc_gdd,
    crop_name = str_to_upper(gdd_temp$crop),
    t_base = gdd_temp$t_base,
    t_upper = gdd_temp$t_upper,
    SIMPLIFY = FALSE) %>% 
    bind_rows()

  saveRDS(final_result, file = paste0('data/weather_results/crop_county_mean_GDD_2000_2023.rds'))

# check data distribution by states
#ggplot(final_result, aes(x = mean_GDD)) +
#    geom_histogram(
#        aes(y = after_stat(density)), 
#        binwidth = 10,  # Adjust based on your data range
#        fill = "steelblue", 
#        color = "black"
#    ) +
#    facet_wrap(~state_alpha, scales = "free_y")

# check data distribution by crops
#ggplot(final_result, aes(x = mean_GDD)) +
#    geom_histogram(
#        aes(y = after_stat(density)), 
#        binwidth = 10,  # Adjust based on your data range
#        fill = "steelblue", 
#        color = "black"
#    ) +
#    facet_wrap(~crop, scales = "free_y")


```

# combine ppt, temp, and GDD

```{r combine}

cum_ppt <- readRDS(file = 'data/weather_results/crop_county_mean_cumulative_ppt_2000_2023.rds')
mean_tmean <- readRDS(file = 'data/weather_results/crop_county_mean_tmean_2000_2023.rds')
mean_GDD <- readRDS(file = 'data/weather_results/crop_county_mean_GDD_2000_2023.rds')

df_list <- list(cum_ppt, mean_tmean, mean_GDD)

weather <- df_list %>%
  reduce(left_join, by = c('state_alpha','GEOID','crop','year'))

saveRDS(weather, file = 'data/intermediate_data/all_crop_county_annual_and_mean_weather_2000_2023.rds')

```
