---
title: "Fertilizer use data processing"
output: html_document
date: "2024-01-29"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
library(tigris)
library(stringr)
```

# read crop data
```{r crop}
crop <- read.csv('data/nass_processed_yield/all_10_crops_county_list_2000_2023.csv')
crop$GEOID <- formatC(crop$GEOID, width = 5, format = 'd' ,flag = '0')

```

# 

```{r fips}
# first match Ninput data names with GEOID from fips code
data(fips_codes) #ML: this is a dataset from the package tigris
fips_codes
fips_codes <- fips_codes %>%
  mutate(GEOID = paste(state_code,county_code, sep = "")) %>%
  mutate(county = str_remove(county, " County")) %>%
  mutate(county = str_remove(county, ' Parish')) %>% # to remove wired names in LA
  mutate(county = case_when(GEOID == '48123' ~  'De Witt',
                            GEOID == '19141' ~ 'O Brien',
                            GEOID == '24033' ~ 'Prince Georges',
                            GEOID == '24035' ~ 'Queen Annes',
                            GEOID == '24037' ~ 'St. Marys',
                            GEOID == '29186' ~ 'Ste Genevieve',
                            .default = county)) %>%
  mutate(county_upper = str_to_upper(county))

```

# read corn N input data: county names have many mismatches 
```{r corn}
# read crop specific N data: corn and winter wheat only, soybean is not available
corn_Ninput_raw <- read.csv('data/raw_data/N_input_data/Corn_county_Ninput_NUE_Nsurplus.csv')

corn_Ninput <- corn_Ninput_raw %>%
  filter(YEAR >= 2000) %>%
  select(YEAR, State,County, Ninput..kg.N.ha.yr.) %>%
  rename(Ninput_kgN.ha.yr = Ninput..kg.N.ha.yr.) %>%
  mutate(state = str_to_title(State)) %>%
  mutate(County = str_replace(County, 'SAINT ','ST. ' )) %>%
  mutate(County = str_replace(County,'^ST ','ST. ' )) %>% # convert the word start with ST to ST.
  mutate(County = case_when(County %in% c('LA MOURE','LA SALLE','LA PORTE') ~ str_replace(County,'^LA ', 'LA'),
                            County %in% c('DE KALB','DE SOTO') ~ str_replace(County,'^DE ', 'DE'), 
                            County == 'STE. GENEVIEVE' ~ 'STE GENEVIEVE', # this name is not consistent in the dataset
                            .default = County)) %>%
  inner_join(fips_codes, by = c('state' = 'state_name', 'County'='county_upper')) %>%
  rename(year = YEAR) %>%
  select(year, GEOID, Ninput_kgN.ha.yr)

# use GEOID to match with corn_GEOID
corn_Ninput_mean <- corn_Ninput %>%
  group_by(GEOID) %>%
  summarise(Ninput_kgN.ha.yr_mean = mean(Ninput_kgN.ha.yr)) %>% # get the mean across years
  filter(GEOID %in% crop$GEOID[crop[['CORN']] == 1]) 

write_rds(corn_Ninput_mean, file = 'data/fertilizer_data/corn_Ninput_county_mean.rds')

# check name differences
corn_diff <- data.frame('GEOID' = setdiff(crop$GEOID[crop[['CORN']] == 1],unique(corn_Ninput_mean$GEOID))) %>%
  left_join(fips_codes, by = 'GEOID') 
  
# 0 counties
```

# read winter_wheat N input data
```{r winter_wheat}
# read crop specific N data: winter_wheat and winter wheat only, soybean is not available
winter_wheat_Ninput_raw <- read.csv('data/raw_data/N_input_data/Winterwheat_county_Ninput_NUE_Nsurplus.csv')

winter_wheat_Ninput <- winter_wheat_Ninput_raw %>%
  filter(YEAR >= 2000) %>%
  select(YEAR, State,County, Ninput..kg.N.ha.yr.) %>%
  rename(Ninput_kgN.ha.yr = Ninput..kg.N.ha.yr.) %>%
  mutate(state = str_to_title(State)) %>%
  mutate(County = str_replace(County, 'SAINT ','ST. ' )) %>%
  mutate(County = str_replace(County,'^ST ','ST. ' )) %>% # convert the word start with ST to ST.
  mutate(County = case_when(County %in% c('LA MOURE','LA SALLE','LA PORTE') ~ str_replace(County,'^LA ', 'LA'),
                            County %in% c('DE KALB','DE SOTO') ~ str_replace(County,'^DE ', 'DE'), 
                            County == 'STE. GENEVIEVE' ~ 'STE GENEVIEVE', # this name is not consistent in the dataset
                            .default = County)) %>%
  inner_join(fips_codes, by = c('state' = 'state_name', 'County'='county_upper')) %>%
  rename(year = YEAR) %>%
  select(year, GEOID, Ninput_kgN.ha.yr)

winter_wheat_Ninput_mean <- winter_wheat_Ninput %>%
  group_by(GEOID) %>%
  summarise(Ninput_kgN.ha.yr_mean = mean(Ninput_kgN.ha.yr)) %>%
  filter(GEOID %in% crop$GEOID[crop[['WHEAT']] == 1]) 

# check name differences
wheat_diff <- data.frame('GEOID' = setdiff(crop$GEOID[crop[['WHEAT']] == 1],unique(winter_wheat_Ninput_mean$GEOID))) %>%
  left_join(fips_codes, by = 'GEOID') 

write_rds(winter_wheat_Ninput_mean, file = 'data/fertilizer_data/wheat_Ninput_county_mean.rds')

```

# read cotton N input data
```{r winter_wheat}
# read crop specific N data: 
cotton_Ninput_raw <- read.csv('data/raw_data/N_input_data/cotton_county_Ninput_NUE_Nsurplus.csv')

cotton_Ninput <- cotton_Ninput_raw %>%
  filter(YEAR >= 2000) %>%
  select(YEAR, State,County, Ninput..kg.N.ha.yr.) %>%
  rename(Ninput_kgN.ha.yr = Ninput..kg.N.ha.yr.) %>%
  mutate(state = str_to_title(State)) %>%
  mutate(County = str_replace(County, 'SAINT ','ST. ' )) %>%
  mutate(County = str_replace(County,'^ST ','ST. ' )) %>% # convert the word start with ST to ST.
  #remove space at the beginning of the string
  mutate(County = case_when(County %in% c('LA MOURE','LA SALLE','LA PORTE') ~ str_replace(County,'^LA ', 'LA'),
                            County %in% c('DE KALB','DE SOTO') ~ str_replace(County,'^DE ', 'DE'), 
                            County == 'LAPAZ' ~ 'LA PAZ', 
                            County == 'STE. GENEVIEVE' ~ 'STE GENEVIEVE', # this name is not consistent in the dataset
                            .default = County)) %>%
  inner_join(fips_codes, by = c('state' = 'state_name', 'County'='county_upper')) %>%
  rename(year = YEAR) %>%
  select(year, GEOID, Ninput_kgN.ha.yr)

cotton_Ninput_mean <- cotton_Ninput %>%
  group_by(GEOID) %>%
  summarise(Ninput_kgN.ha.yr_mean = mean(Ninput_kgN.ha.yr)) %>%
  filter(GEOID %in% crop$GEOID[crop[['COTTON']] == 1]) 

# check name differences
cotton_diff <- data.frame('GEOID' = setdiff(crop$GEOID[crop[['COTTON']] == 1],unique(cotton_Ninput_mean$GEOID))) %>%
  left_join(fips_codes, by = 'GEOID') 

write_rds(cotton_Ninput_mean, file = 'data/fertilizer_data/cotton_Ninput_county_mean.rds')

```

# read barley N input data
```{r winter_wheat}
# read crop specific N data: 
barley_Ninput_raw <- read.csv('data/raw_data/N_input_data/Barley_county_Ninput_NUE_Nsurplus.csv')

barley_Ninput <- barley_Ninput_raw %>%
  filter(YEAR >= 2000) %>%
  select(YEAR, State,County, Ninput..kg.N.ha.yr.) %>%
  rename(Ninput_kgN.ha.yr = Ninput..kg.N.ha.yr.) %>%
  mutate(state = str_to_title(State)) %>%
  mutate(County = str_replace(County, 'SAINT ','ST. ' )) %>%
  mutate(County = str_replace(County,'^ST ','ST. ' )) %>% # convert the word start with ST to ST.
  #remove space at the beginning of the string
  mutate(County = case_when(County %in% c('LA MOURE','LA SALLE','LA PORTE') ~ str_replace(County,'^LA ', 'LA'),
                            County %in% c('DE KALB','DE SOTO') ~ str_replace(County,'^DE ', 'DE'), 
                            County == 'LAPAZ' ~ 'LA PAZ', 
                            County == 'STE. GENEVIEVE' ~ 'STE GENEVIEVE', # this name is not consistent in the dataset
                            .default = County)) %>%
  inner_join(fips_codes, by = c('state' = 'state_name', 'County'='county_upper')) %>%
  rename(year = YEAR) %>%
  select(year, GEOID, Ninput_kgN.ha.yr)

barley_Ninput_mean <- barley_Ninput %>%
  group_by(GEOID) %>%
  summarise(Ninput_kgN.ha.yr_mean = mean(Ninput_kgN.ha.yr)) %>%
  filter(GEOID %in% crop$GEOID[crop[['BARLEY']] == 1]) 

# check name differences
barley_diff <- data.frame('GEOID' = setdiff(crop$GEOID[crop[['BARLEY']] == 1],unique(barley_Ninput_mean$GEOID))) %>%
  left_join(fips_codes, by = 'GEOID') 

write_rds(barley_Ninput_mean, file = 'data/fertilizer_data/barley_Ninput_county_mean.rds')

```

# read sorghum N input data
```{r winter_wheat}
# read crop specific N data: 
sorghum_Ninput_raw <- read.csv('data/raw_data/N_input_data/sorghum_county_Ninput_NUE_Nsurplus.csv')

sorghum_Ninput <- sorghum_Ninput_raw %>%
  filter(YEAR >= 2000) %>%
  select(YEAR, State,County, Ninput..kg.N.ha.yr.) %>%
  rename(Ninput_kgN.ha.yr = Ninput..kg.N.ha.yr.) %>%
  mutate(state = str_to_title(State)) %>%
  mutate(County = str_replace(County, 'SAINT ','ST. ' )) %>%
  mutate(County = str_replace(County,'^ST ','ST. ' )) %>% # convert the word start with ST to ST.
  #remove space at the beginning of the string
  mutate(County = case_when(County %in% c('LA MOURE','LA SALLE','LA PORTE') ~ str_replace(County,'^LA ', 'LA'),
                            County %in% c('DE KALB','DE SOTO') ~ str_replace(County,'^DE ', 'DE'), 
                            County == 'LAPAZ' ~ 'LA PAZ', 
                            County == 'STE. GENEVIEVE' ~ 'STE GENEVIEVE', # this name is not consistent in the dataset
                            .default = County)) %>%
  inner_join(fips_codes, by = c('state' = 'state_name', 'County'='county_upper')) %>%
  rename(year = YEAR) %>%
  select(year, GEOID, Ninput_kgN.ha.yr)

sorghum_Ninput_mean <- sorghum_Ninput %>%
  group_by(GEOID) %>%
  summarise(Ninput_kgN.ha.yr_mean = mean(Ninput_kgN.ha.yr)) %>%
  filter(GEOID %in% crop$GEOID[crop[['SORGHUM']] == 1]) 

# check name differences
sorghum_diff <- data.frame('GEOID' = setdiff(crop$GEOID[crop[['SORGHUM']] == 1],unique(sorghum_Ninput_mean$GEOID))) %>%
  left_join(fips_codes, by = 'GEOID') 

write_rds(sorghum_Ninput_mean, file = 'data/fertilizer_data/sorghum_Ninput_county_mean.rds')

```

# read NuGIS data for oats, tobacco, and sugarbeets
```{r data}
# read NuGIS data: county-level nutrient data (N, P2O5, K2O) averaged across crops
nugis_raw <- read.csv('data/raw_data/N_input_data/NuGIS_County_Data.csv')

nugis <- nugis_raw %>%
  filter(year >= 2000) %>%
  select(year, FIPS, FarmLbsN_perAc, FarmLbsP2O5_perAc, FarmLbsK2O_perAc)
nugis$GEOID <- formatC(nugis$FIPS, width = 5, format = 'd' ,flag = '0')

# oats
nugis_n_mean_oats <- nugis %>%
  group_by(GEOID) %>%
  summarise(FarmLbsN_perAc_mean = mean(FarmLbsN_perAc),
            FarmLbsP2O5_perAc_mean = mean(FarmLbsP2O5_perAc),
            FarmLbsK2O_perAc_mean = mean(FarmLbsK2O_perAc)) %>%
  filter(GEOID %in% crop$GEOID[crop[['OATS']] == 1]) 

write_rds(nugis_n_mean_oats, file = 'data/fertilizer_data/oats_Ninput_county_mean.rds')

# tobacco
nugis_n_mean_tobacco <- nugis %>%
  group_by(GEOID) %>%
  summarise(FarmLbsN_perAc_mean = mean(FarmLbsN_perAc),
            FarmLbsP2O5_perAc_mean = mean(FarmLbsP2O5_perAc),
            FarmLbsK2O_perAc_mean = mean(FarmLbsK2O_perAc)) %>%
  filter(GEOID %in% crop$GEOID[crop[['TOBACCO']] == 1]) 

write_rds(nugis_n_mean_tobacco, file = 'data/fertilizer_data/tobacco_Ninput_county_mean.rds')

# sugarbeets
nugis_n_mean_sugarbeets <- nugis %>%
  group_by(GEOID) %>%
  summarise(FarmLbsN_perAc_mean = mean(FarmLbsN_perAc),
            FarmLbsP2O5_perAc_mean = mean(FarmLbsP2O5_perAc),
            FarmLbsK2O_perAc_mean = mean(FarmLbsK2O_perAc)) %>%
  filter(GEOID %in% crop$GEOID[crop[['SUGARBEETS']] == 1]) 

write_rds(nugis_n_mean_sugarbeets, file = 'data/fertilizer_data/sugarbeets_Ninput_county_mean.rds')

```

Not RUN!
# check the correlation between two data sources for corn
```{r check}
nugis_n_mean_corn <- nugis %>%
  group_by(GEOID) %>%
  summarise(FarmLbsN_perAc_mean = mean(FarmLbsN_perAc),
            FarmLbsP2O5_perAc_mean = mean(FarmLbsP2O5_perAc),
            FarmLbsK2O_perAc_mean = mean(FarmLbsK2O_perAc)) %>%
  filter(GEOID %in% crop$GEOID[crop[['CORN']] == 1]) 

#write.csv(nugis_n_mean_corn, file = 'data/AFT-data/Fertilizer-data/corn_county_NuGIS_county_mean.csv')


check_n <- nugis_n_mean_corn %>%
  full_join(corn_Ninput_mean, by = 'GEOID') %>%
  na.omit()

ggplot(data = check_n) +
  geom_point(aes(x=FarmLbsN_perAc_mean, y = Ninput_kgN.ha.yr_mean))

cor(check_n$FarmLbsN_perAc_mean, check_n$Ninput_kgN.ha.yr_mean)
# 0.3069056

ggplot(data = check_n) +
  geom_point(aes(x=FarmLbsN_perAc_mean, y = FarmLbsP2O5_perAc_mean))

ggplot(data = check_n) +
  geom_point(aes(x=FarmLbsN_perAc_mean, y = FarmLbsK2O_perAc_mean))


# check the map
library(usmap)
library(colorspace)
# get state boundaries
states <- us_map(regions = 'states', exclude = c('AK','HI'))

# get county boundaries 
county <- us_map(regions = 'counties')

# combine data
Nugis_map <- nugis_n_mean %>%
  left_join(county, by = c('GEOID'='fips'))

corn_N_map <- corn_Ninput_mean %>%
  left_join(county, by = c('GEOID'='fips'))

# all data is matched
#colnames(corn_na_map) <- 'GEOID'
#corn_na_map <- corn_na_map %>%
#  left_join(county, by = c('GEOID'='fips'))

#create the map
ggplot() +
  geom_polygon(data = Nugis_map, aes(x =x, y=y, group = group,fill=FarmLbsN_perAc_mean)) +
  geom_polygon(data = states, aes(x = x, y = y, group = group), fill = NA, color = 'black') +
  scale_fill_continuous_divergingx(palette = 'Spectral', mid = 0, n_interp = 11, name = 'FarmLbsN_perAc_mean') + 
  labs(title = 'Corn') +
  coord_equal() +
  theme_void() 

#ggplot() +
#  geom_polygon(data = corn_N_map, aes(x =x, y=y, group = group,fill=Ninput_kgN.ha.yr_mean)) +
#  geom_polygon(data = corn_na_map, aes(x =x, y=y, group = group),fill = NA, color='red') +
#  geom_polygon(data = states, aes(x = x, y = y, group = group), fill = NA, color = 'black') +
#  scale_fill_continuous_divergingx(palette = 'Spectral', mid = 0, n_interp = 11, name = 'Ninput_kgN.ha.yr_mean') + 
#  labs(title = 'Corn') +
#  coord_equal() +
#  theme_void() 

ggplot() +
  geom_polygon(data = Nugis_map, aes(x =x, y=y, group = group,fill=FarmLbsP2O5_perAc_mean)) +
  geom_polygon(data = states, aes(x = x, y = y, group = group), fill = NA, color = 'black') +
  scale_fill_continuous_divergingx(palette = 'Spectral', mid = 0, n_interp = 11, name = 'FarmLbsP2O5_perAc_mean') + 
  labs(title = 'Corn') +
  coord_equal() +
  theme_void() 

```

# check the correlation between two data sources for winter_wheat
```{r check}
nugis_n_mean_winter_wheat <- nugis %>%
  group_by(GEOID) %>%
  summarise(FarmLbsN_perAc_mean = mean(FarmLbsN_perAc),
            FarmLbsP2O5_perAc_mean = mean(FarmLbsP2O5_perAc),
            FarmLbsK2O_perAc_mean = mean(FarmLbsK2O_perAc)) %>%
  filter(GEOID %in% crop$GEOID[crop[['WHEAT']] == 1]) 

#write.csv(nugis_n_mean_winter_wheat, file = 'data/AFT-data/Fertilizer-data/wheat_county_NuGIS_county_mean.csv')


check_n_winter_wheat <- nugis_n_mean_winter_wheat %>%
  full_join(winter_wheat_Ninput_mean, by = 'GEOID') %>%
  na.omit()

ggplot(data = check_n_winter_wheat) +
  geom_point(aes(x=FarmLbsN_perAc_mean, y = Ninput_kgN.ha.yr_mean))

cor(check_n_winter_wheat$FarmLbsN_perAc_mean, check_n_winter_wheat$Ninput_kgN.ha.yr_mean)

ggplot(data = check_n_winter_wheat) +
  geom_point(aes(x=FarmLbsN_perAc_mean, y = FarmLbsP2O5_perAc_mean))

ggplot(data = check_n_winter_wheat) +
  geom_point(aes(x=Ninput_kgN.ha.yr_mean, y = FarmLbsP2O5_perAc_mean))

# check the map
library(usmap)
library(colorspace)
# get state boundaries
states <- us_map(regions = 'states', exclude = c('AK','HI'))

# get county boundaries 
county <- us_map(regions = 'counties')

# combine data
Nugis_map_winter_wheat <- nugis_n_mean_winter_wheat %>%
  left_join(county, by = c('GEOID'='fips'))

winter_wheat_N_map <- winter_wheat_Ninput_mean %>%
  left_join(county, by = c('GEOID'='fips'))

#create the map
ggplot() +
  geom_polygon(data = Nugis_map_winter_wheat, aes(x =x, y=y, group = group,fill=FarmLbsN_perAc_mean)) +
  geom_polygon(data = states, aes(x = x, y = y, group = group), fill = NA, color = 'black') +
  scale_fill_continuous_divergingx(palette = 'Spectral', mid = 0, n_interp = 11, name = 'FarmLbsN_perAc_mean') + 
  labs(title = 'winter_wheat') +
  coord_equal() +
  theme_void() 

ggplot() +
  geom_polygon(data = winter_wheat_N_map, aes(x =x, y=y, group = group,fill=Ninput_kgN.ha.yr_mean)) +
  geom_polygon(data = states, aes(x = x, y = y, group = group), fill = NA, color = 'black') +
  scale_fill_continuous_divergingx(palette = 'Spectral', mid = 0, n_interp = 11, name = 'Ninput_kgN.ha.yr_mean') + 
  labs(title = 'winter_wheat') +
  coord_equal() +
  theme_void() 

ggplot() +
  geom_polygon(data = Nugis_map, aes(x =x, y=y, group = group,fill=FarmLbsP2O5_perAc_mean)) +
  geom_polygon(data = states, aes(x = x, y = y, group = group), fill = NA, color = 'black') +
  scale_fill_continuous_divergingx(palette = 'Spectral', mid = 0, n_interp = 11, name = 'FarmLbsP2O5_perAc_mean') + 
  labs(title = 'winter_wheat') +
  coord_equal() +
  theme_void() 

```

# check data sources for oats
```{r check}

# check the map
library(usmap)
library(colorspace)
# get state boundaries
states <- us_map(regions = 'states', exclude = c('AK','HI'))

# get county boundaries 
county <- us_map(regions = 'counties')

# combine data
Nugis_map_oats <- nugis_n_mean_oats %>%
  left_join(county, by = c('GEOID'='fips'))

winter_wheat_N_map <- winter_wheat_Ninput_mean %>%
  left_join(county, by = c('GEOID'='fips'))

#create the map
ggplot() +
  geom_polygon(data = Nugis_map_oats, aes(x =x, y=y, group = group,fill=FarmLbsN_perAc_mean)) +
  geom_polygon(data = states, aes(x = x, y = y, group = group), fill = NA, color = 'black') +
  scale_fill_continuous_divergingx(palette = 'Spectral', mid = 0, n_interp = 11, name = 'FarmLbsN_perAc_mean') + 
  labs(title = 'oats') +
  coord_equal() +
  theme_void() 

ggplot() +
  geom_polygon(data = Nugis_map_oats, aes(x =x, y=y, group = group,fill=FarmLbsP2O5_perAc_mean)) +
  geom_polygon(data = states, aes(x = x, y = y, group = group), fill = NA, color = 'black') +
  scale_fill_continuous_divergingx(palette = 'Spectral', mid = 0, n_interp = 11, name = 'FarmLbsP2O5_perAc_mean') + 
  labs(title = 'winter_wheat') +
  coord_equal() +
  theme_void() 

```